---
title: "PCA"
author: "Hannah Damico"
date: "2022-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(factoextra)
library(data.table)
library(tidyverse)
```


```{r}
train <- read.csv("train_recode.csv")
test <- read.csv("test_recode.csv")

data <- rbind(train, test)
```




1. Explore dataset

Show first 5 rows of data
```{r}
head(data,5)
```

Get variable names of dataset
```{r}
# names(data)
```

Find columns with missing data
```{r}
# Get number of NAs per variable
na_table  = sapply(data, function(y) sum(length(which(is.na(y)))))

# Filter table to only include variables with NA
na_table = na_table[na_table > 0]
# na_table

# Get columns with missing data
missing_cols = names(na_table)
# missing_cols

# Remove Extra Index Variable X
missing_cols = missing_cols[-1]

# Remove outcome from missing cols
missing_cols = missing_cols[-length(missing_cols)]

# missing_cols
```

Mean imputation of missing values
```{r warning = FALSE}
# Copy dataset
clean_data = data

# Iterate over missing columns
# for(missing_col in missing_cols) {
#   # Mean imputation for survival group
#   clean_data[[missing_col]][is.na(data[[missing_col]]) & data$outcome == 0] =
#     mean(data[[missing_col]][data$outcome == 0], na.rm = TRUE)
#   
#   # Mean imputation for non-survival group
#   clean_data[[missing_col]][is.na(data[[missing_col]]) & data$outcome == 1] =
#     mean(data[[missing_col]][data$outcome == 1], na.rm = TRUE)
# }

clean_data <- data %>% 
  group_by(subjid) %>% 
  tidyr::fill(missing_cols, .direction = c("updown"))


# Remove row with NA outcome
clean_data = clean_data[!is.na(clean_data$In.hospital_death),]
```

Show that no NAs remain
```{r}
# Get number of NAs per variable
na_table_new  = sapply(clean_data, function(y) sum(length(which(is.na(y)))))

na_table_new
```

Extract only continuous variables
```{r}
# De-select non-continuous variables
cont_data = clean_data %>% 
  select(-contains(c("MechVent", "Gender", "ICUType")), - X)
```

Show first 5 rows of continuous data
```{r}
head(cont_data)
```
Show summary of data
```{r}
summary(cont_data)
```
Get dimensions of data
```{r}
nrow(cont_data)
ncol(cont_data)
```
There are 1815 observations and 154 continuous variables.

2. It is important to standardize the variables to have mean zero and standard deviation one before performing PCA.

Calculate PCs
```{r message = FALSE}
cont_data <- cont_data %>%  mutate_all(~replace(., is.na(.), -1))
pr.out=prcomp(cont_data, scale=TRUE)
```

prcomp() function centers the variables to have mean zero. By using the option scale=TRUE, we scale the variables to have standard deviation one. 

Print output categories of pc object
```{r}
names(pr.out)
```
Print variable means
```{r}
pr.out$center

```

```{r}
pr.out$scale^2
```

3. The rotation matrix provides the principal component loadings; each col- umn of pr.out$rotation contains the corresponding principal component loading vector.

I only outputted the first 5 variables for the first 2 PCs because there are a lot of PCs.

```{r}
pr.out$rotation[,1:2]
```

```{r}
ncol(pr.out$rotation)
```
There are 38 principal components because the maximum number of principal component loadings is the minimum between (sample size - 1) and number of predictors. We have 38 predictors and 1176 individuals in our sample, so the maximum number of principal components is 38.

4. The 1176 × 38 matrix x has as its columns the principal component score vectors. That is, the kth column is the kth principal component score vector.

```{r}
dim(pr.out$x)
```

```{r}
pr.out$x[1:20,1:2]
```
I only printed out the first 20 observations for the first two components for simplicity sake.

4. Plot the first 2 principal components.

```{r, fig.width = 5, fig.height = 5}
fviz_pca_biplot(pr.out, repel = TRUE, label = "var", select.var = list(contrib = 4), arrowsize = 0.75, alpha.ind = 0.1, col.var = "blue",title = "PCA Biplot with top 4 contributing variables")
```

5. The prcomp() function also outputs the standard deviation of each principal component. For instance, on the data set, we can access these standard deviations as follows:

```{r}
pr.var=pr.out$sdev^2
pr.var
```

Proportion of variance explained.

```{r}
pve=pr.var/sum(pr.var)
```


```{r}
plot(pve, xlab="Principal Component", ylab="Proportion of Variance Explained ", ylim=c(0,1),type='b')
```

```{r}
plot(cumsum(pve), xlab="Principal Component ", ylab=" Cumulative Proportion of Variance Explained ", ylim=c(0,1), type='b')
```


